---
title: "Lab 5 : Data wrangling and graphing COVID-19 reporting data (Part 2)"
output: html_document
---

Transform Tables:
https://rstudio-education.github.io/tidyverse-cookbook/transform-tables.html

## Exercises

#### 1. Revise the above example for joining tables using 6/13/2020 and 9/13/2020. Plot the data as a bar plot. Make sure your graph looks nice in the report by adjusting the height in the R code chunk header.

```{r}
library(tidyverse)
```

```{r}
Confirmed_State_6_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/06-13-2020.csv")) %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 
str(Confirmed_State_6_13)
```
```{r}
Confirmed_State_9_13 <-   read_csv(url("https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_daily_reports/09-13-2020.csv")) %>% 
  filter (Country_Region == "US") %>% 
  group_by(Province_State, Country_Region) %>% 
  summarise(Confirmed = sum(Confirmed)) 
str(Confirmed_State_9_13)
```
```{r}
# check whether any different
setdiff(Confirmed_State_9_13$Province_State, Confirmed_State_6_13$Province_State)
```
Since all of the states and territories in 9/13/2020 are found in 6/13/2020 a left_join or full_join can be used to merge the tables
```{r}
# join tables
Confirmed_State_6_13_9_13_joined <- full_join(Confirmed_State_6_13,
      Confirmed_State_9_13, by = c("Province_State"))
head(Confirmed_State_6_13_9_13_joined)
```

```{r}
tail(Confirmed_State_6_13_9_13_joined, 5)
```

```{r}
# check whether any NA exist
which(is.na(Confirmed_State_6_13_9_13_joined))
```
```{r}
Confirmed_State_6_13_9_13_joined <- full_join(Confirmed_State_6_13,
      Confirmed_State_9_13, by = c("Province_State")) %>% 
      rename(Confirmed_6_13_2020 = "Confirmed.x", Confirmed_9_13_2020 = "Confirmed.y") %>% 
      select(-Country_Region.x, -Country_Region.y)
      #replace_na(list(Confirmed_3_13_2020 = 0))
head(Confirmed_State_6_13_9_13_joined)
```

### Switching between wide and long table formats.

gather and spead were used in the Data Carpentries tutorial to switch between wide and long table formats. They work fine, but are being replaced by the newer functions pivot_longer and pivot_wider. which we will use below. See the pivot vignette for other examples.

```{r}
Confirmed_State_6_13_9_13_joined_long <- Confirmed_State_6_13_9_13_joined %>% 
              pivot_longer(-c(Province_State),
                            names_to = "Date", values_to = "Confirmed")
str(Confirmed_State_6_13_9_13_joined_long)
```

```{r, fig.width = 10, fig.height = 10}
ggplot(Confirmed_State_6_13_9_13_joined_long, aes(x = Confirmed,  y = Province_State))  + 
    geom_point(aes(color = Date)) 
```
#### 2. Add more informative x and y labels and a title to the above graph in Ex2.

```{r, fig.width = 10, fig.height = 10}
tem_plot <-ggplot(Confirmed_State_6_13_9_13_joined_long, aes(x = Confirmed,  y = Province_State))  + 
    geom_point(aes(color = Date))
tem_plot +
  labs(x = "Confirmed number", y = "States of US") +
  labs(title = "Confirmed number Plot ", subtitle = "Time: 06132020 & 09132020") +
  labs(caption = "(based on data from https://github.com/CSSEGISandData/COVID-19)")
  
```

### Working with the time series data

#### 3. Using the time series data. Make a plot of the total number of confirmed deaths per day worldwide

Confirmed:
```{r}
#load data
download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv", 
               destfile = "data/time_series_covid19_confirmed_global.csv")
```
```{r}
time_series_confirmed <- read_csv("data/time_series_covid19_confirmed_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
head(time_series_confirmed)
```


```{r}
time_series_confirmed_long <- time_series_confirmed %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Confirmed") 
head(time_series_confirmed_long)
```

Death:
```{r}
download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv", 
               destfile = "data/time_series_covid19_deaths_global.csv")
```

```{r}
time_series_deaths <- read_csv("data/time_series_covid19_deaths_global.csv")%>%
  rename(Province_State = "Province/State", Country_Region = "Country/Region")
```


```{r}
time_series_deaths_long <- time_series_deaths %>% 
               pivot_longer(-c(Province_State, Country_Region, Lat, Long),
                            names_to = "Date", values_to = "Deaths") 
head(time_series_deaths_long)
```

##### Joining the time series tables

To join 2 tables we need a common column in which each row is a unique name. This is often called the Key. None of the columns in this data set meet that criteria, but we can create a column to serve as our key by mergering the names for the Province_State, Country_Region and Date columns using unite

```{r}
time_series_confirmed_long <- time_series_confirmed_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".", remove = FALSE)
head(time_series_confirmed_long)
```


```{r}
time_series_deaths_long <- time_series_deaths_long %>% 
  unite(Key, Province_State, Country_Region, Date, sep = ".") %>% 
  select(Key, Deaths)
head(time_series_deaths_long)
```
Now to join the tables. A full_join which means that the keys that are common to the tables will be joined and there will be counts for both Confirmed and Deaths. Where there are not matching values, returns NA for the one missing. select can be used to remove the key after the join since we don’t have further need for it.

```{r}
time_series_long_joined <- full_join(time_series_confirmed_long,
              time_series_deaths_long, by = c("Key")) %>% 
              select(-Key)
head(time_series_long_joined)
```
```{r}
# NA Check
which(is.na(time_series_long_joined$Confirmed))
```
```{r}
which(is.na(time_series_long_joined$Deaths))
```
There are no NA values. Now we are in good shape to proceed to using these table for graphing, but first reformat Date

```{r}
library(lubridate)
```

```{r}
#transofmr time data from format 1/22/20 to 2020-01-22
time_series_long_joined$Date <- mdy(time_series_long_joined$Date)
```
Confirmed and Deaths are count data and there may be times when we want to plot both on the same graph. There are multiple ways to accomplish this and one is through creating a column using _pivot_long

```{r}
time_series_long_joined_counts <- time_series_long_joined %>% 
  pivot_longer(-c(Province_State, Country_Region, Lat, Long, Date),
               names_to = "Report_Type", values_to = "Counts")
head(time_series_long_joined_counts)
```

##### Making Graphs from the time series data
To make a times series graph of the confirmed cases we need to summarize the Country date to count up the individual state data for the US.

```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Confirmed", "Deaths"), sum) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Deaths")
```
```{r}
time_series_long_joined %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Confirmed", "Deaths"), sum) %>% 
  filter (Country_Region %in% c("China","Japan", "Korea, South",
                                "Italy","Spain", "US")) %>% 
    ggplot(aes(x = Date,  y = Deaths)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths") +
    facet_wrap(~Country_Region, ncol=2, scales="free_y")
```
Now several countries on the same graph
```{r}
time_series_long_joined %>% 
    group_by(Country_Region,Date) %>% 
    summarise_at(c("Confirmed", "Deaths"), sum) %>% 
    filter (Country_Region %in% c("China","France","Italy", 
                                "Korea, South", "US")) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Deaths")
```
We can use the alternative data frame with column Report_Type to show Confirmed and Deaths. Because the counts for Deaths are low relative to Confirmed a log scale works best here for seeing the pattern


```{r}
time_series_long_joined_counts %>% 
  group_by(Country_Region, Report_Type, Date) %>% 
  summarise(Counts = sum(Counts)) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = log2(Counts), fill = Report_Type, color = Report_Type)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Cases")
```

#### 4. Use Mutate to make a new column with deaths/confirmed cases (Hint: this might be best done after joining the tables, but before pivoting longer).

```{r}
time_series_long_joined_Mutated <- time_series_long_joined %>% 
  mutate(Deaths_Confirmed_Cases = Confirmed + Deaths)
head(time_series_long_joined_Mutated)
```

#### 5. Plot US deaths/confirmed cases per day.

```{r}
time_series_long_joined_Mutated %>% 
  group_by(Country_Region,Date) %>% 
  summarise_at(c("Deaths_Confirmed_Cases"), sum) %>% 
  filter (Country_Region == "US") %>% 
    ggplot(aes(x = Date,  y = Deaths_Confirmed_Cases)) + 
    geom_point() +
    geom_line() +
    ggtitle("US COVID-19 Deaths/Confirmed Cases")

```

#### 6. Make a single graph with from the times series data with the 10 countries with the hightest death total.

```{r}
top_ten_deaths_countrys <- time_series_long_joined_Mutated %>% 
  group_by(Country_Region) %>% 
  summarise_at(c("Deaths"), sum) %>% 
  arrange(desc(Deaths)) %>%
  slice(1:10) 
head(top_ten_deaths_countrys)
```

```{r}
time_series_long_joined_Mutated %>% 
  group_by(Country_Region, Date) %>% 
  summarise_at(c("Confirmed", "Deaths","Deaths_Confirmed_Cases"), sum) %>% 
  filter(Country_Region %in% top_ten_deaths_countrys$Country_Region) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("Worldwide top 10 COVID-19 Deaths")
```

#### 7. Using Facet_Wraps make a graph from the times series data with the 10 countries with the hightest death total. Format the table to look nice in the report by changing the number of rows in your graph and/or adjusting the height in the R code chunk header.

```{r, fig.width = 10, fig.height = 10}
time_series_long_joined_Mutated %>% 
  group_by(Country_Region, Date) %>% 
  summarise_at(c("Confirmed", "Deaths","Deaths_Confirmed_Cases"), sum) %>% 
  filter(Country_Region %in% top_ten_deaths_countrys$Country_Region) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("Worldwide top 10 COVID-19 Deaths") +
  facet_wrap(~Country_Region, ncol=2, scales="free_y")
```

#### 8. Using Facet_Wraps make a graph from the times series data for confirmed cases for all US states and territories. Format the table to look nice in the report by changing the number of rows in your graph and/or adjusting the height in the R code chunk header.


```{r}
#load data
download.file(url="https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_US.csv", 
               destfile = "data/time_series_covid19_confirmed_US.csv")
```
```{r}
time_series_confirmed_US <- read_csv("data/time_series_covid19_confirmed_US.csv")
head(time_series_confirmed_US)
```
```{r}
# Just keep the useful information:
time_series_confirmed_US_Reorganized <- time_series_confirmed_US %>%
  filter(Country_Region == 'US') %>% 
  select(-UID, -iso2, -iso3, -code3, -FIPS, -Admin2, -Lat, -Long_, -Combined_Key) %>% 
  group_by()

```

```{r}
time_series_confiremd_US_long <- time_series_confirmed_US_Reorganized %>% 
               pivot_longer(-c(Province_State, Country_Region),
                            names_to = "Date", values_to = "Confirmed") 
time_series_confiremd_US_long$Date <- mdy(time_series_confiremd_US_long$Date)
head(time_series_confiremd_US_long)
```
```{r}
which(is.na(time_series_confiremd_US_long$Confirmed))
```

```{r, fig.width = 8, fig.height = 50}
time_series_confiremd_US_long %>% 
    group_by(Province_State,Date) %>% 
    summarise_at(c("Confirmed"), sum) %>% 
    ggplot(aes(x = Date,  y = Confirmed, color = Province_State)) + 
    geom_point() +
    geom_line() +
    ggtitle("COVID-19 Confirmed for each US") +
   facet_wrap(~Province_State, ncol=2, scales="free_y") +
  theme(legend.position = 'bottom')
```

#### 9. Apply a theme other than the one used in the Data Carpentries tutorial to the graph in Ex8.

```{r, fig.width = 10, fig.height = 10}
time_series_long_joined_Mutated %>% 
  group_by(Country_Region, Date) %>% 
  summarise_at(c("Confirmed", "Deaths","Deaths_Confirmed_Cases"), sum) %>% 
  filter(Country_Region %in% top_ten_deaths_countrys$Country_Region) %>% 
    ggplot(aes(x = Date,  y = Deaths, color = Country_Region)) + 
    geom_point() +
    geom_line() +
    ggtitle("Worldwide top 10 COVID-19 Deaths") +
  facet_wrap(~Country_Region, ncol=2, scales="free_y") +
  theme_bw()
```


