---
title: "Lab 9A : Programming in R - Regular Expressions"
output: html_document
---

https://jeffreyblanchard.github.io/EvoGeno/EvoGeno_Lab9_stringr.html#exercises

https://jrnold.github.io/r4ds-exercise-solutions/strings.html

## Regular Expressions
“A regular expression (shortened as regex or regexp) is a sequence of characters that define a search pattern. Usually such patterns are used by string-searching algorithms for”find" or “find and replace” operations on strings, or for input validation. It is a technique developed in theoretical computer science and formal language theory.

The concept arose in the 1950s when the American mathematician Stephen Cole Kleene formalized the description of a regular language. The concept came into common use with Unix text-processing utilities. Different syntaxes for writing regular expressions have existed since the 1980s, one being the POSIX standard and another, widely used, being the Perl syntax.

Regular expressions are used in search engines, search and replace dialogs of word processors and text editors, in text processing utilities such as sed and AWK and in lexical analysis. Many programming languages provide regex capabilities either built-in or via libraries." - Wikipedia

## Base R Functions

R’s regular expression utilities work similar as in other languages. To learn how to use them in R, one can consult the main help page on this topic with ?regexp. Here is an overview from a Basic Regular Expressions in R Cheat Sheet. https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf

The Base R functions for dealing with regular expressions are

1. grep(), grepl(): These functions search for matches of a regular expression/pattern in a character vector. grep() returns the indices into the character vector that contain a match or the specific strings that happen to have the match. grepl() returns a TRUE/FALSE vector indicating which elements of the character vector contain a match
2. regexpr(), gregexpr(): Search a character vector for regular expression matches and return the indices of the string where the match begins and the length of the match
3. sub(), gsub(): Search a character vector for regular expression matches and replace that match with another string
4. regexec(): This function searches a character vector for a regular expression, much like regexpr(), but it will additionally return the locations of any parenthesized sub-expressions. Probably easier to explain through demonstration.

## The stringr Package

The stringr package is part of the tidyverse collection of packages and wraps they underlying stringi package in a series of convenience functions. Some of the complexity of using the base R regular expression functions is usefully hidden by the stringr functions. In addition, the stringr functions provide a more rational interface to regular expressions with more consistency in the arguments and argument ordering.

Given what we have discussed so far, there is a fairly straightforward mapping from the base R functions to the stringr functions. In general, for the stringr functions, the data are the first argument and the regular expression is the second argument, with optional arguments afterwards.

str_subset() is much like grep(value = TRUE) and returns a character vector of strings that contain a given match.

str_detect() is essentially equivalent grepl().

str_extract() plays the role of regexpr() and regmatches(), extracting the matches from the output.

Finally, str_match() does the job of regexec() by provide a matrix containing the parenthesized sub-expressions.

### String manipulation with stringr

The below text is shortened version of the in the Strings Strings Chapter in R for Data Science by Hadley Wickham. For quick tips see

  stringr and regulalar expression cheat sheet https://stringr.tidyverse.org/
  stringr tidyverse page https://stringr.tidyverse.org/
  
## 14.3 Matching patterns with regular expressions

Regexps are a very terse language that allow you to describe patterns in strings. They take a little while to get your head around, but once you understand them, you’ll find them extremely useful.

To learn regular expressions, we’ll use str_view() and str_view_all(). These functions take a character vector and a regular expression, and show you how they match. We’ll start with very simple regular expressions and then gradually get more and more complicated. Once you’ve mastered pattern matching, you’ll learn how to apply those ideas with various stringr functions.

```{r}
library(stringr)
library(tidyverse)
```

### 14.3.1 Basic matches

The simplest patterns match exact strings:

```{r}
x <- c("apple", "banana", "pear")
str_view(x, "an")
```

The next step up in complexity is ., which matches any character (except a newline):

```{r}
str_view(x, ".a.")
```

Note that this only shows the first match in a string.

But if “.” matches any character, how do you match the character “.”? You need to use an “escape” to tell the regular expression you want to match it exactly, not use its special behaviour. Like strings, regexps use the backslash, \, to escape special behaviour. So to match an ., you need the regexp \.. Unfortunately this creates a problem. We use strings to represent regular expressions, and \ is also used as an escape symbol in strings. So to create the regular expression \. we need the string "\\.".

```{r}
# To create the regular expression, we need \\
dot <- "\\."

# But the expression itself only contains one:
writeLines(dot)
## \.
```
```{r}
# And this tells R to look for an explicit .
x <- c("abc", "a.c", "bef")
str_view(x, "a\\.c")
```

If \is used as an escape character in regular expressions, how do you match a literal \? Well you need to escape it, creating the regular expression \\. To create that regular expression, you need to use a string, which also needs to escape \. That means to match a literal \ you need to write "\\\\" — you need four backslashes to match one!

```{r}
x <- "a\\b"
writeLines(x)
```

```{r}
str_view(x, "\\\\")
```

In this book, I’ll write regular expression as . and strings that represent the regular expression as "\\.".

#### 14.3.1.1 Exercises

1. Explain why each of these strings don’t match a \: "\", "\\", "\\\".
Becase the program think you want to use it as a special behaviour, not want to match it exactly

2. How would you match the sequence "'\?

```{r}
E12 <- c("abc", "a.'\\c", "bef")
```

```{r}
str_view(E12, "'\\\\")
```

3. What patterns will the regular expression \..\..\.. match? How would you represent it as a string?

```{r}
E13 <- c("abc", "a\\..\\..\\..c", "bef")
writeLines(E13)
```

```{r}
str_view(E13, "\\\\..\\\\..\\\\..")
```

### 14.3.2 Anchors

By default, regular expressions will match any part of a string. It’s often useful to anchor the regular expression so that it matches from the start or end of the string. You can use:

  ^ to match the start of the string
  $ to match the end of the string.

```{r}
x <- c("apple", "banana", "pear")
str_view(x, "^a")
```

```{r}
str_view(x, "a$")
```

To remember which is which, try this mnemonic which I learned from Evan Misshula: if you begin with power (^), you end up with money ($).

To force a regular expression to only match a complete string, anchor it with both ^ and $:

```{r}
x <- c("apple pie", "apple", "apple cake")
str_view(x, "apple")
```

```{r}
str_view(x, "^apple$")
```

You can also match the boundary between words with \b. I don’t often use this in R, but I will sometimes use it when I’m doing a search in RStudio when I want to find the name of a function that’s a component of other functions. For example, I’ll search for \bsum\b to avoid matching summarise, summary, rowsum and so on.

#### 14.3.2.1 Exercises

1. How would you match the literal string "$^$"?

```{r}
E21 <- c("abc", "a$^$c", "bef")
writeLines(E21)
```

```{r}
str_view(E21, "\\$\\^\\$")
```

2. Given the corpus of common words in stringr::words, create regular expressions that find all words that:

```{r}
E22 <- c(stringr::words)
```

  Start with “y”.
  
```{r}
str_view(E22, "^y", match = TRUE)
```
  
  End with “x”
  
```{r}
str_view(E22, "x$", match = TRUE)
```
  
  Are exactly three letters long. (Don’t cheat by using str_length()!)
  
```{r}
str_view(E22, "\\b[a-zA-Z0-9]{3}\\b", match = TRUE)
```
  
  Have seven letters or more.

```{r}
str_view(E22, "\\b[a-zA-Z0-9]{7,}\\b", match = TRUE)
```

### 14.3.3 Character classes and alternatives

There are a number of special patterns that match more than one character. You’ve already seen ., which matches any character apart from a newline. There are four other useful tools:

1. \d : matches any digit.
2. \s : matches any whitespace (e.g. space, tab, newline).
3. [abc] : matches a, b, or c.
4. [^abc] : matches anything except a, b, or c.

Remember, to create a regular expression containing \d or \s, you’ll need to escape the \ for the string, so you’ll type "\\d" or "\\s".

A character class containing a single character is a nice alternative to backslash escapes when you want to include a single metacharacter in a regex. Many people find this more readable.

```{r}
# Look for a literal character that normally has special meaning in a regex
x <- c("abc", "a.c", "a*c", "a c")
str_view(x, "a[.]c")
```

```{r}
str_view(x, ".[*]c")
```

```{r}
str_view(x, "a[ ]")
```

This works for most (but not all) regex metacharacters: $ . | ? * + ( ) [ {. Unfortunately, a few characters have special meaning even inside a character class and must be handled with backslash escapes: ] \ ^ and -.

You can use alternation to pick between one or more alternative patterns. For example, abc|d..f will match either ‘“abc”’, or “deaf”. Note that the precedence for | is low, so that abc|xyz matches abc or xyz not abcyz or abxyz. Like with mathematical expressions, if precedence ever gets confusing, use parentheses to make it clear what you want:

```{r}
x <- c("grey", "gray")
str_view(x, "gr(e|a)y")
```

#### 14.3.3.1 Exercises

Create regular expressions to find all words that:
  
  Start with a vowel.
  
```{r}
str_view(stringr::words, "^[aeiou]", match = TRUE)
```
  
  That only contain consonants. (Hint: thinking about matching “not”-vowels.)
  
```{r}
str_view(stringr::words, "[aeiou]", match=FALSE)
```
  
  End with ed, but not with eed.
  
```{r}
str_view(stringr::words, "[^e]ed$", match = TRUE)
```
  
```{r}
str_view(c("ed", stringr::words), "(^|[^e])ed$", match = TRUE)
```
  
  End with ing or ise.
  
```{r}
str_view(stringr::words, "(ing|ise)$", match=TRUE)
```
  
  Is “q” always followed by a “u”?
  
```{r}
str_view(stringr::words, "q[^u]", match = TRUE)
```
  
### 14.3.4 Repetition

The next step up in power involves controlling how many times a pattern matches:

  ?: 0 or 1
  +: 1 or more
  *: 0 or more

```{r}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"
str_view(x, "CC?")
```

```{r}
str_view(x, "CC+")
```

```{r}
str_view(x, 'C[LX]+')
```

Note that the precedence of these operators is high, so you can write: colou?r to match either American or British spellings. That means most uses will need parentheses, like bana(na)+

You can also specify the number of matches precisely:

  {n}: exactly n
  {n,}: n or more
  {,m}: at most m
  {n,m}: between n and m
  
```{r}
str_view(x, "C{2}")
```
  
```{r}
str_view(x, "C{2,}")
```

By default these matches are “greedy”: they will match the longest string possible. You can make them “lazy”, matching the shortest string possible by putting a ? after them. This is an advanced feature of regular expressions, but it’s useful to know that it exists:

```{r}
str_view(x, 'C{2,3}?')
```

```{r}
str_view(x, 'C[LX]+?')
```

#### 14.3.4.1 Exercises

1. Describe the equivalents of ?, +, * in {m,n} form.

? == {0,1}
+ == {1,}
* == {0,}

2. Describe in words what these regular expressions match: (read carefully to see if I’m using a regular expression or a string that defines a regular expression.)

  ^.*$ : Any string
  "\\{.+\\}" : {} include at least one character
  \d{4}-\d{2}-\d{2} : number combination like 1234-56-78
  "\\\\{4}": \\{4} == \\\\

3. Create regular expressions to find all words that:

  Start with three consonants.
```{r}
str_view(stringr::words, "^[^aeiou]{3}", match=TRUE)
```
  
  Have three or more vowels in a row.

```{r}
str_view(stringr::words, "[aeiou]{3,}", match=TRUE)
```
 
  
  Have two or more vowel-consonant pairs in a row.
  
```{r}
str_view(stringr::words, "([aeiou][^aeiou]){2,}", match=TRUE)
```
  
### 14.3.5 Grouping and backreferences

Earlier, you learned about parentheses as a way to disambiguate complex expressions. Parentheses also create a numbered capturing group (number 1, 2 etc.). A capturing group stores the part of the string matched by the part of the regular expression inside the parentheses. You can refer to the same text as previously matched by a capturing group with backreferences, like \1, \2 etc. For example, the following regular expression finds all fruits that have a repeated pair of letters.

(Shortly, you’ll also see how they’re useful in conjunction with str_match().)

```{r}
str_view(fruit, "(..)\\1", match = TRUE)
```

#### 14.3.5.1 Exercises

1. Describe, in words, what these expressions will match:

(.)\1\1 : character repeat three times
"(.)(.)\\2\\1" : a pair of character repeated by reversed order, like abba
(..)\1 : two character, repeat one time
"(.).\\1.\\1" : original character, other character, original character, other character, original character, like "abaca", "b8b.b"
"(.)(.)(.).*\\3\\2\\1" : Three characters followed by zero or more characters of any kind followed by the same three characters but in reverse order. E.g. "abcsgasgddsadgsdgcba" or "abccba" or "abc1cba"

2. Construct regular expressions to match words that:

Start and end with the same character.
```{r}
str_view(stringr::words, "^(.)((.*\\1$)|\\1?$)", match=TRUE)
```

Contain a repeated pair of letters (e.g. “church” contains “ch” repeated twice.)

```{r}
str_view(stringr::words, "([A-Za-z][A-Za-z]).*\\1", match=TRUE)
```

Contain one letter repeated in at least three places (e.g. “eleven” contains three “e”s.)

```{r}
str_view(stringr::words, "([a-z]).*\\1.*\\1", match=TRUE)
```

